---
title: "Data Analysis NOAA Storm Database"
author: "Afsah"
date: "03/06/2020"
output: html_document
---

## This analysis attempts to answer two questions using the NOAA Storm Database. The first is: what events in the US are most harmful to population health? The second is: what types of events have the greatest economic consequences in the US?

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading libraries

```{r}
library(knitr)
library(ggplot2)
```

## Processing the data

```{r}
if(!file.exists("stormData.csv.bz2")) {
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
  destfile = "stormData.csv.bz2", method = "curl")
}

# Loading data
dsNOAA <- read.csv(bzfile("stormData.csv.bz2"), sep=",", header=T)
head(dsNOAA)
```

### Subsetting the database

```{r}
tidyNOAA <- dsNOAA[,c('EVTYPE','FATALITIES','INJURIES', 'PROPDMG', 'PROPDMGEXP', 'CROPDMG', 'CROPDMGEXP')]
head(tidyNOAA)
```

```{r}
str(tidyNOAA)
```

When calculating the economic damage, the variables as follow should be calculated:
PROPDMG and CROPDMG (which mean property damage and crop damage)

## Converting H, K, M, B units to compute PROPDMG

### Creating an empty column

```{r}
tidyNOAA$PROPDMGNUM = 0
```

### Filling in the data with proper units

```{r}
tidyNOAA[tidyNOAA$PROPDMGEXP == "H", ]$PROPDMGNUM = tidyNOAA[tidyNOAA$PROPDMGEXP == "H", ]$PROPDMG * 10^2
tidyNOAA[tidyNOAA$PROPDMGEXP == "K", ]$PROPDMGNUM = tidyNOAA[tidyNOAA$PROPDMGEXP == "K", ]$PROPDMG * 10^3
tidyNOAA[tidyNOAA$PROPDMGEXP == "M", ]$PROPDMGNUM = tidyNOAA[tidyNOAA$PROPDMGEXP == "M", ]$PROPDMG * 10^6
tidyNOAA[tidyNOAA$PROPDMGEXP == "B", ]$PROPDMGNUM = tidyNOAA[tidyNOAA$PROPDMGEXP == "B", ]$PROPDMG * 10^9

head(tidyNOAA, 100)
```

## Converting H, K, M, B units to compute CROPDMG

### Creating an empty column

```{r}
tidyNOAA$CROPDMGNUM = 0
```

### Assigning proper values based off parameters

```{r}
tidyNOAA[tidyNOAA$CROPDMGEXP == "H", ]$CROPDMGNUM = tidyNOAA[tidyNOAA$CROPDMGEXP == "H", ]$CROPDMG * 10^2
tidyNOAA[tidyNOAA$CROPDMGEXP == "K", ]$CROPDMGNUM = tidyNOAA[tidyNOAA$CROPDMGEXP == "K", ]$CROPDMG * 10^3
tidyNOAA[tidyNOAA$CROPDMGEXP == "M", ]$CROPDMGNUM = tidyNOAA[tidyNOAA$CROPDMGEXP == "M", ]$CROPDMG * 10^6
tidyNOAA[tidyNOAA$CROPDMGEXP == "B", ]$CROPDMGNUM = tidyNOAA[tidyNOAA$CROPDMGEXP == "B", ]$CROPDMG * 10^9
```

```{r}
head(tidyNOAA,100)
```

## Answering questions

## Quesion 1: Across the US, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

### Plotting number of fatalities with most harmful event type

```{r}
fatalities <- aggregate(FATALITIES ~ EVTYPE, data=tidyNOAA, sum)

fatalities <- fatalities[order(-fatalities$FATALITIES), ][1:10, ]
fatalities$EVTYPE <- factor(fatalities$EVTYPE, levels = fatalities$EVTYPE)

ggplot(fatalities, aes(x = EVTYPE, y = FATALITIES)) + 
    geom_bar(stat = "identity", fill = "blue", las = 3) + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
    xlab("Event Type") + ylab("Fatalities") + ggtitle("Number of fatalities by top 10 Weather Events")
```

### Plotting number of injuries according to most harmful even type

```{r}
injuries <- aggregate(INJURIES ~ EVTYPE, data=tidyNOAA, sum)
injuries <- injuries[order(-injuries$INJURIES), ][1:10, ]
injuries$EVTYPE <- factor(injuries$EVTYPE, levels = injuries$EVTYPE)

ggplot(injuries, aes(x = EVTYPE, y = INJURIES)) + 
    geom_bar(stat = "identity", fill = "blue", las = 3) + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
    xlab("Event Type") + ylab("Injuries") + ggtitle("Number of injuries by top 10 Weather Events")
```

## Conclusion for Question 1: Tornadoes cause most harm. According to the plots, they cause most fatalities and injuries.

## Question 2: Across the United States, which types of events hae the greatest economic consequences?

### Plotting number of damages according to most harmful event type

```{r}
damages <- aggregate(PROPDMGNUM + CROPDMGNUM ~ EVTYPE, data=tidyNOAA, sum)
names(damages) = c("EVTYPE", "TOTALDAMAGE")
damages <- damages[order(-damages$TOTALDAMAGE), ][1:10, ]
damages$EVTYPE <- factor(damages$EVTYPE, levels = damages$EVTYPE)

ggplot(damages, aes(x = EVTYPE, y = TOTALDAMAGE)) + 
    geom_bar(stat = "identity", fill = "blue", las = 3) + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
    xlab("Event Type") + ylab("Damages ($)") + ggtitle("Property & Crop Damages by top 10 Weather Events")
```


### Conclusion for Question 2: Floods cause the most economic tragedies.